Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

### 2021-04-14 analysis/paper/paper.Rmd
REMOVE REFERENCES TO STABLE STAGE
stable stage distributions and application to haul-out corrections are being moved to a separate paper for the Western Bering.

In this study, we use data from 15 years of bio-logger records in a comprehensive
investigation of haul-out behavior by 3 ice-associated seals in the Bering and Chukchi
Seas. We focused our effort on bearded, spotted, and ribbon seals and do not include
ringed seals in the analysis. Ringed seals are unique in their use of snow lairs in the
spring months, and they warrant a separate study that includes additional environmental
covariates, such as the onset of snow melt and snow depth. Our goals are threefold. First,
we wish to establish baseline estimates for the chronology of haul-out behavior in the
critical spring season for each species across different age and sex classes. Second, we
wish to refine estimates of haul-out availability corrections for aerial surveys in order
to improve estimates of seal abundance. Previously estimated availability correction
factors (e.g., [@bengtson2005pb], [@conn2014mee], and [@verhoef2014sm]) accounted for
variables such as the time-of-day and day-of-year, but did not investigate meteorological
variables that have been shown to influence haul-out behavior of walruses [@reder2003pb;
@udevitz2009cjz]. Further, previous correction factor analyses assumed that tagged seals
were representative of the population as a whole, even though the age and sex distribution
of telemetered individuals might differ from the population. Third, we explore the annual
variability in haul-out timing and possible linkage to dramatic changes in the
availability and extent of seasonal sea ice between 2005 and 2020. Our work extends the
scope of previous haul-out analyses, includes the influence of meteorological variability
and possible ramifications of non-representative age-sex samples, and investigates the
potential impact of a changing icescape on the behavior of these species.

@@

## Stable stage distributions

::: {custom-style="comment"}
I didn't update this at all from Paul's previous text other than to revise some of the
latex equations and add references. Would be good to confirm all of this is still
appropriate and makes sense.
:::

Previous availability correction factors used in aerial surveys for ice-associated Arctic
seals (e.g. [@bengtson2005pb], [@conn2014mee], and [@verhoef2014sm]) ignored any age- and
sex-based variation in haul-out probabilities. The implicit assumption is that the
behavior of telemetered animals represents the population as a whole. However, if haul-out
probabilities are related to sex- and age-class, and if there are tangible differences
between the age- and sex-structure of the telemetered sample and that of the population,
correction factors computed in this way may lead to biased abundance estimates.

To determine how much of a potential problem this is within our data, we compared
estimates of haul-out probabilities that (1) ignored sex- and age-structure and (2)
accounted for sex- and age-structure, adjusting for the relative proportion of seals in
each sex- and age-class. To simplify comparisons, we fitted two additional GLMPMs to
ribbon and spotted seal haul-out datasets. In the first model, we expressed haul-out
probabilities as a function of solar hour and day-of-year. In the second model, we also
included effects of sex-age class and sex-age interactions with solar hour and
day-of-year. To adjust for the proportion of seals in each age class, we calculated a
predicted population-averaged haul-out probability at solar noon as a function of
day-of-year ($t$) as

$$
\widetilde{p_t}=\sum_{s}\sum_{a}{\omega_{s,a}{\hat{p}}_{s,a,t}}
$$

where $\omega_{s,a}$ is the proportion of the population of seals in age class $a$ and sex
$s$. Typically, we do not know $\omega_{s,a}$ exactly, but can use information about
likely mortality and reproductive schedules of seals to approximate $\omega_{s,a}$ using
stable stage distributions calculated from matrix population models ().

To calculate plausible values for $\omega_{s,a}$, we obtained natural mortality
predictions from a hierarchical meta-analysis of phocid natural mortality
[@trukhanova2018ee]. We then assumed a 50/50 sex ratio at birth, 1 pup per litter, and
reproductive schedules as reported by Fedoseev [-@fedoseev2000] for seals in the western
Bering Sea to formulate Leslie matrices Ai for each species i. Stable age distributions
$v_i={v\_i0,v\_i1,⋯,v\_iA}$ were then determined by normalizing the dominant eigenvector
of $A_i$ so that they summed to one . Given a stable age distribution, we calculated the
proportion of the population by sex and stage (young-of-the-year: \<1 year; subadult: at
least one year but sexually immature; adult: sexually mature) using data on sexual
maturity from Fedoseev [-@fedoseev2000]. In particular, the expected proportion of the
population that are sex $s$ and young-of-year was set as $\pi_{i,s,yoy}=0.5v_i0$, while
the expected proportion that are subadults and adults were set to

$$
\pi_{i,s,sub}={0.5\sum_{a}{{(1-m}_{i,s,a})}v}_{ia}
$$

and

$$
\pi_{i,s,ad}={0.5\sum_{a}\ m_{i,s,a}v}_{ia}
$$

Here, $m_{i,s,a}$ gives the anticipated proportion of animals of species $i$ and sex $s$
that are sexually mature. Further details on mortality, reproductive and maturity
schedules, and matrix modeling methods are presented in Appendix A.

@@

## Stable stage distributions

::: {custom-style="comment"}
This section is unchanged from Paul's original write up. I have put off revisiting and
integrating with my current document/code. Maybe between Paul/Irina/Me we can sort out a
quick update. Might just need to walk me through the previous code and make sure I get it
properly incorporated with this version
:::

Stable stage distributions calculated using mortality and reproductive schedules produced
different patterns for the four species of ice-associated seals residing in Alaskan waters
(Fig. X). Ribbon seals mature the fastest, and so can be expected to have the largest
proportion of adults and smallest proportion of subadults; conversely, ringed seals mature
slowly and are predicted to have roughly equal proportions of adults and subadults (note
that we do not attempt to model ringed seal haul-out behavior in this study; see
Discussion). Population-level haul-out distributions were somewhat different depending on
whether we a) adjusted haul-out probabilities based on stable stage distributions or b)
assumed the tagged sample represented the population. In particular, ribbon seal estimates
differed somewhat at the beginning of the time series, but the two approaches produced
similar estimates (Fig. X) within the window that previous aerial surveys in the region
have been conducted (April and May; e.g. Conn et al. 2014). By contrast, during this time
period spotted seal haul-out estimates were often ≈10% higher for the approach that
adjusted for stable age distributions, presumably because subadults had lower predicted
haul-out probabilities than adults but made up a disproportionately high proportion of the
tagged sample.

@@

Previous attempts to estimate the abundance of phocid seals from aerial survey data in the
Bering and Chukchi Seas (e.g. Bengtson et al. 2005, Conn et al. 2014, Ver Hoef et al.
2014) have used estimated haul-out probabilities to correct for the proportion of animals
that are in the water and thus unavailable to be counted during aerial surveys. Although
several of these studies allowed haul-out probabilities to vary by day-of-year and
time-of-day, previous correction factors have not accounted for variability between years,
different weather conditions, and the age-sex class of the sample. In this paper, we have
shown that there can be considerable differences in the number of seals hauled out on ice
based on these factors. We recommend that future abundance analyses employ correction
factors that account for these factors. For instance, it is relatively straightforward to
obtain weather reanalysis products for times and locations that are surveyed and to
construct a relevant correction factor based on predictions of GMPMs. Further, one can
apply stable stage distribution estimates from matrix population models to adjust
predictions to the "average" prediction of the population. The latter suggestion is
particularly appropriate when the age structure of the tagged sample does not reflect the
likely age structure of the population. For instance, population weighted estimates can
differ substantially from those based on a naïve analysis of the tagged sample, as was the
case with analysis of spotted seal data (e.g. Fig X). The most challenging element in
developing availability correction factors is with annual variability. It can be difficult
to get a sufficient sample size to estimate year-specific correction factors, particularly
because research teams would likely need to tag seals and conduct aerial surveys
concurrently which requires considerably more personnel and money. One possible suggestion
is to estimate a "shift" parameter within models for aerial survey counts that allow the
peak of haul-out distributions to be adjusted earlier or later in the year based on the
frequency of counts observed over time. Regardless, researchers should anticipate there
being some unmodeled heterogeneity in availability probability present in abundance
estimates obtained from aerial surveys. This may make trend detection difficult, as one
will not know if moderate differences in abundance estimates are attributable to changes
in abundance or changes in haul-out frequencies.

>>>>

In this study, we use data from 15 years of bio-logger records in a comprehensive
investigation of haul-out behavior by 3 ice-associated seals in the Bering and Chukchi
Seas. We focused our effort on bearded, spotted, and ribbon seals and do not include
ringed seals in the analysis. Ringed seals are unique in their use of snow lairs in the
spring months, and they warrant a separate study that includes additional environmental
covariates, such as the onset of snow melt and snow depth. Our goals are threefold. First,
we wish to establish baseline estimates for the chronology of haul-out behavior in the
critical spring season for each species across different age and sex classes. Second, we
wish to refine estimates of haul-out availability corrections for aerial surveys in order
to improve estimates of seal abundance. Previously estimated availability correction
factors (e.g., [@bengtson2005pb], [@conn2014mee], and [@verhoef2014sm]) accounted for
variables such as the time-of-day and day-of-year, but did not investigate meteorological
variables that have been shown to influence haul-out behavior of walruses [@reder2003pb;
@udevitz2009cjz]. Third, we explore the annual variability in haul-out timing and possible
linkage to dramatic changes in the availability and extent of seasonal sea ice between
2005 and 2020. Our work extends the scope of previous haul-out analyses, includes the
influence of meteorological variability and possible ramifications of non-representative
age-sex samples, and investigates the potential impact of a changing icescape on the
behavior of these species.

@@

@@

@@

Previous attempts to estimate the abundance of phocid seals from aerial survey data in the
Bering and Chukchi Seas (e.g. Bengtson et al. 2005, Conn et al. 2014, Ver Hoef et al.
2014) have used estimated haul-out probabilities to correct for the proportion of animals
that are in the water and thus unavailable to be counted during aerial surveys. Although
several of these studies allowed haul-out probabilities to vary by day-of-year and
time-of-day, previous correction factors have not accounted for variability between years,
different weather conditions, and the age-sex class of the sample. In this paper, we have
shown that there can be considerable differences in the number of seals hauled out on ice
based on these factors. We recommend that future abundance analyses employ correction
factors that account for these factors. For instance, it is relatively straightforward to
obtain weather reanalysis products for times and locations that are surveyed and to
construct a relevant correction factor based on predictions of GMPMs. The most challenging
element in developing availability correction factors is with annual variability. It can
be difficult to get a sufficient sample size to estimate year-specific correction factors,
particularly because research teams would likely need to tag seals and conduct aerial
surveys concurrently which requires considerably more personnel and money. One possible
suggestion is to estimate a "shift" parameter within models for aerial survey counts that
allow the peak of haul-out distributions to be adjusted earlier or later in the year based
on the frequency of counts observed over time. Regardless, researchers should anticipate
there being some unmodeled heterogeneity in availability probability present in abundance
estimates obtained from aerial surveys. This may make trend detection difficult, as one
will not know if moderate differences in abundance estimates are attributable to changes
in abundance or changes in haul-out frequencies.
